Title: "Reproducible Research: Peer Assignment 1"
Author: "Mohamed Rizwan"
Date: "March 15, 2019"
Output: html_document
##loading libraries
```{r, results = 'hide', echo = FALSE}
library(dplyr)
library(lattice)
```
##Downloaded csv file from coursera in the working directory

#unzip the file to get "activity.csv" file in the working directory
```{r}
unzip("repdata_data_activity.zip")
```
#1.reading and preprocessing the data
```{r,cache= TRUE, echo=TRUE}
act <- read.csv("activity.csv")
act$date <- as.Date(act$date)
head(act)
```
#2.Histogram of the total number of steps taken each day
```{r}
by_date <- act %>% select(date, steps) %>% group_by(date) %>% summarize(totalsteps= sum(steps)) %>% na.omit()
hist(by_date$totalsteps, xlab = "Total number of steps taken each day", main="Histogram of Total number of steps taken each day", breaks = 20)
```

#3.Mean and median of the total number of steps taken per day
```{r}
summary(by_date)
```
##From the summary, Mean = 10766  and Median = 10765

#4.Time series plot of the average number of steps taken
```{r, echo=TRUE, results='show'}
library(ggplot2)
by_interval <- act %>% select(interval, steps) %>% na.omit() %>% group_by(interval) %>% summarize(averagesteps= mean(steps))
ggplot(by_interval, aes(x=interval, y=averagesteps))+ geom_line()
```

#5.The 5-minute interval that, on average, contains the maximum number of steps
```{r}
by_interval[which(by_interval$averagesteps == max(by_interval$averagesteps)),]
```
#6.Code describing for imputing missing data
##Total number of missing values, NAs in the dataset
```{r}
missingvals <- sum(is.na(act))
```
#missingvals
```{r}
missingvals
```
##Replacing the NAs with mean for that 5 minutes interval the day
```{r}
replacewithmean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
meandata <- act %>% group_by(interval) %>% mutate(steps=replacewithmean(steps))
head(meandata)
```
##Histogram of the total number of steps taken each day of the imputed dataset
```{r}
by_date2 <- meandata %>% select(date, steps) %>% group_by(date) %>% summarize(totalsteps= sum(steps))
```

##Mean and median of the total number of steps taken per day
```{r}
summary(by_date2)
```
##From the summary, Mean = 10766  and Median = 10766

#Histogram of the total number of steps taken each day with the NAs imputed/replaced
```{r}
hist(by_date2$totalsteps, xlab = "Total number of steps taken each day", main="Histogram of Total number of steps taken each day", breaks = 20)
```

##Difference in the means
```{r}
mean_1 <- mean(by_date$totalsteps, na.rm = TRUE)
mean_2 <- mean(by_date2$totalsteps)
difference <- mean_2-mean_1
```
#difference
```{r}
difference
```
## There's no difference in the means of the total steps from the previous dataset(by_date) and imputed dataset(by_date2) 

#6.Differences in activity patterns between weekdays and weekends:
```{r, results= 'hide'}
meandata$day <- weekdays(meandata$date)
meandata$weekend <- ifelse(meandata$day=="Saturday" | meandata$day=="Sunday", "Weekend", "Weekday")
```
##Mean of the steps on the weekdays and weekends
```{r}
mean_weekend_weekday <- aggregate(meandata$steps, by= list(meandata$weekend, meandata$interval), na.omit(mean))
names(mean_weekend_weekday) <- c("weekend", "interval", "steps")
head(mean_weekend_weekday)
```
##Panel plot showing the average number of steps in each interval weekdays and weekends
```{r, results="show"}
ggplot(mean_weekend_weekday, aes(x= interval, y= steps, color=weekend)) + geom_line()+
    facet_grid(weekend ~.) + xlab("5 min-Interval") + ylab("Average number of the Steps taken") +
    ggtitle("Comparison of average number of steps in each interval")
```
